On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-11-23 manuscripts/PNAS/second review/resub/chazthm_pnas_resub.tex
new xi

The constant $C_2 = C_2(\mathbf{A}, \mathcal{H})$, in turn, will be presented relative to a quantity from \cite{Deutsch12} used to analyze the convergence of the alternating projections algorithm. 
More specifically, in Lem.~\ref{DistanceToIntersectionLemma} below, the following definition is used to bound the distance between a point and the intersection of subspaces given an upper bound on its distance from each individual subspace.

\begin{definition}\label{SpecialSupportSet}\label{FriedrichsDefinition}
For a collection of real subspaces $\mathcal{V} = \{V_i\}_{i=1}^\ell$, define $\xi := 1$ when $|\mathcal{V}| = 1$ and otherwise:
\begin{align*}
\xi(\mathcal{V}) := 1 - \left(1 -  \max \prod_{i=1}^{\ell-1} \sin^2  \theta \left(V_i, \cap_{j>i} V_j \right)  \right)^{1/2},
\end{align*} 
%
where the maximum is taken over all orderings 
%\footnote{We modify the quantity in \cite{Deutsch12} in this way since the subspace ordering is irrelevant to our purpose.} 
of the $V_i$ and the angle $\theta \in (0,\frac{\pi}{2}]$ is defined implicitly as \cite[Def.~9.4]{Deutsch12}:
\begin{align*}
\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \substack{ \mathbf{u} \in U \cap (U \cap W)^\perp, \ \|\mathbf{u}\|_2 \leq 1 \\ \mathbf{w} \in W \cap (U \cap W)^\perp, \  \|\mathbf{w}\|_2 \leq 1 } \right\}.
\end{align*}
\end{definition}
Note that $\theta \in (0,\frac{\pi}{2}]$ implies $0 < \xi \leq 1$, and that $\xi(\mathcal{V}') \geq \xi(\mathcal{V})$ when $\mathcal{V}' \subseteq \mathcal{V}$.\footnote{We acknowledge the counter-intuitive property that $\theta =  \pi/2$ when $U \subseteq W$.}  
The constant $C_2$ in Lem.~\ref{MainLemma} can then be expressed as:  
\begin{align}\label{Cdef2}
	C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \in {\mathcal{H} \choose r+1}} \xi( \bm{\mathcal{A}}_\mathcal{G} ) },
\end{align}
% C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \in {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1}} \xi( \{ \bm{\mathcal{A}}_S \}
which we remark is consistent with the assumption on $C_1$ in the proof of the case $k=1$ at the beginning of this section.

@@

It follows from $L_{2\mathcal{H}}(\mathbf{A}) > 0$ and the general linear position of the $\mathbf{x}_i$ that $L(\mathbf{AX}_{K}) > 0$; that is, the columns of the $n \times k$ matrix $\mathbf{AX}_K$ form a basis for $\bm{\mathcal{A}}_S$. Fixing $\mathbf{0} \neq \mathbf{y} \in \bm{\mathcal{A}}_S$, there then exists $\mathbf{0} \neq \mathbf{c} = (c_1, \ldots, c_k) \in \mathbb{R}^k$ such that $\mathbf{y} = \mathbf{AX}_K\mathbf{c}$. Setting \mbox{$\mathbf{\overline{y}} = \mathbf{B\overline{X}}_K\mathbf{c} \in \bm{\mathcal{B}}_{\overline S}$}, we have:
\begin{align*}
\|\mathbf{y} - \mathbf{\overline{y}}\|_2 
&= \|\sum_{i=1}^k c_i(\mathbf{AX}_K - \mathbf{B\overline{X}}_K)_i\|_2
\leq \varepsilon \sum_{i=1}^k |c_i| \\
&\leq \varepsilon \sqrt{k}  \|\mathbf{c}\|_2 
\leq \frac{\varepsilon}{L(\mathbf{AX}_K)} \|\mathbf{y}\|_2,
\end{align*}
where the last inequality follows directly from \eqref{Ldef}. From Def.~\ref{dDef}:
\begin{align}\label{rhs222}
d(\bm{\mathcal{A}}_S, \bm{\mathcal{B}}_{\overline S}) 
\leq \frac{\varepsilon}{  L(\mathbf{AX}_{K}) } \leq \varepsilon \frac{C_1}{C_2},
\end{align}
%
where the second inequality is due to $L(\mathbf{AX}_{K}) \geq L_k(\mathbf{AX}_{I(S)})$ and \eqref{Cdef1}. Finally, apply Lem.~\ref{MainLemma} with $\varepsilon < L_2(\mathbf{A})/C_1$. % L_2(A) > 0 by def of L, since L_2H(A) > 0 and H covers [m]
\end{proof}

@@

\begin{lemma}\label{DistanceToIntersectionLemma}
Fix $k \geq 2$. Let $\mathcal{V} = \{V_i\}_{i=1}^k$ be a collection of subspaces of $\mathbb{R}^m$ and set $V = \cap_{i = 1}^k V_i$. For  $\mathbf{x} \in \mathbb{R}^m$, we have (where $\xi$ is given in Def.~\ref{SpecialSupportSet}):
\begin{align}\label{DTILeq}
\text{\rm dist}(\mathbf{x}, V) \leq \frac{1}{\xi(\mathcal{V})} \sum_{i=1}^k \text{\rm dist}(\mathbf{x}, V_i).
\end{align}
\end{lemma}
\vspace{-.3 cm}
\begin{proof} 
Recall the orthogonal projection onto a subspace $V \subseteq \mathbb{R}^m$ is the mapping $\Pi_V: \mathbb{R}^m \to V$ that associates with each $\mathbf{x}$ its unique nearest point in $V$; i.e., $\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 = \text{\rm dist}(\mathbf{x}, V)$.
Next, observe:
\begin{align}\label{f}
\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 &\leq \|\mathbf{x} - \Pi_{V_k} \mathbf{x}\|_2 + \|\Pi_{V_k}  \mathbf{x} - \Pi_{V_k}\Pi_{V_{k-1}}\mathbf{x}\|_2 \nonumber \\
&\ \ \ + \cdots + \|\Pi_{V_k} \Pi_{V_{k-1}}\cdots \Pi_{V_1} \mathbf{x} - \Pi_V \mathbf{x}\|_2 \nonumber \\
&\leq \|\Pi_{V_k}\cdots\Pi_{V_{1}} \mathbf{x} - \Pi_V \mathbf{x}\|_2 + \sum_{\ell=1}^k \|\mathbf{x} - \Pi_{V_{\ell}} \mathbf{x}\|_2,
\end{align}
using the triangle inequality and that the spectral norm satisfies $\|\Pi_{V_{\ell}}\|_2 \leq 1$ for all $\ell$ (since $\Pi_{V_{\ell}}$ are orthogonal projections).

@@

Finally, substituting this into \eqref{f} and rearranging produces \eqref{DTILeq} after replacing $1 - z$ with $\xi(\mathcal{V})$.
\end{proof}
\begin{lemma}\label{NonEmptyLemma} 
Fix an $r$-regular hypergraph $\mathcal{H} \subseteq 2^{[m]}$ satisfying the SIP. If the map $\pi: \mathcal{H} \to 2^{[\overline m]}$ has $\sum_{S \in \mathcal{H}} |\pi(S)| \geq \sum_{S \in \mathcal{H}} |S|$ and:
\begin{align}\label{cond}
	|\cap \pi(\mathcal{G})| \leq |\cap \mathcal{G} |,\ \ \   \text{for } \mathcal{G} \in {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1},
\end{align}
%
then $\overline m \geq m$, and if $\overline m < mr / (r-1)$ then the map $i \mapsto \cap \pi(\sigma(i))$ is an injective map to $[\overline m]$ from some $J \subseteq [m]$ of size $\overline m - r(\overline m - m)$. %In particular, if $\overline m = m$ then $\pi$ is induced by a permutation on $[m]$.
\end{lemma}
\begin{proof}
Consider the set: $T_1 := \{(i, S): i \in \pi(S), S \in \mathcal{H}\}$, which number $|T_1| = \sum_{S \in \mathcal{H}} |\pi(S)| \geq \sum_{S \in \mathcal{H}} |S| = \sum_{i \in [m]} \deg_\mathcal{H}(i) = mr$. Note that \eqref{cond} implies $|T_1| \leq \overline m r$, since otherwise pigeonholing the elements of $T_1$ with respect to their possible first indices $[\overline m]$ would imply $\deg_\mathcal{H}(i) > r$ for some $i \in [m]$. Thus, $\overline m \geq m$.

@@

We next show that \eqref{cond} holds.  Fix $\mathcal{G} \subseteq {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1}$. Since $\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})} \subseteq \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)}$, if $\cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)} = \{\textbf{0}\}$, then we must have $|\cap \pi(\mathcal{G})| = 0$ (as the columns of $\mathbf{B}_{\pi(S)}$ are linearly independent) and \eqref{cond} is trivially true. Suppose then that the intersection is not zero. Keeping in mind that $d(U',V) \leq d(U,V)$ whenever $U' \subseteq U$ and applying (in order) Lem.~\ref{spanIntersectionLemma}, Lem.~\ref{DistanceToIntersectionLemma}, \eqref{eqdim}, \eqref{GapUpperBound}, and \eqref{Cdef2} gives:
\begin{align}\label{randoml}
d(\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})}&, \bm{\mathcal{A}}_{\cap \mathcal{G}}  ) 
\leq d\left( \cap \bm{\mathcal{B}}_{\pi(\mathcal{G})}, \cap \bm{\mathcal{A}}_\mathcal{G} \right)
\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \cap \bm{\mathcal{B}}_{\pi(\mathcal{G})},\bm{\mathcal{A}_{T}} \right) }{ \xi( \bm{\mathcal{A}}_\mathcal{G} ) } \nonumber \\
&\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \bm{\mathcal{B}}_{\pi(T)},\bm{\mathbf{A}}_{T} \right) }{ \xi( \bm{\mathcal{A}}_\mathcal{G} ) }
\leq \frac{|\mathcal{G}| \varepsilon}{\xi( \bm{\mathcal{A}}_\mathcal{G} )} 
\leq \frac{C_2 \varepsilon}{\max_j\|\mathbf{A}_j\|_2}. 
\end{align}

>>>>

The constant $C_2 = C_2(\mathbf{A}, \mathcal{H})$, in turn, will be presented relative to a quantity from \cite{Deutsch12} used to analyze the convergence of the alternating projections algorithm. 
Specifically, in Lem.~\ref{DistanceToIntersectionLemma} below, the following definition is used to bound the distance between a point and the intersection of subspaces given an upper bound on its distance from each individual subspace.

\begin{definition}\label{SpecialSupportSet}\label{FriedrichsDefinition}
For a collection of real subspaces $\mathcal{V} = \{V_i\}_{i=1}^\ell$, define $\xi^2 := 1$ when $|\mathcal{V}| = 1$, and otherwise:
\begin{align}\label{xi}
\xi^2(\mathcal{V}) := 1 -  \max \prod_{i=1}^{\ell-1} \sin^2  \theta \left(V_i, \cap_{j>i} V_j \right) ,
\end{align} 
%
where the maximum is taken over all orderings 
%\footnote{We modify the quantity in \cite{Deutsch12} in this way since the subspace ordering is irrelevant to our purpose.} 
of the $V_i$ and the angle $\theta \in (0,\frac{\pi}{2}]$ is defined implicitly as \cite[Def.~9.4]{Deutsch12}:
\begin{align*}
\cos{\theta(U,W)} := \max\left\{ |\langle \mathbf{u}, \mathbf{w} \rangle|: \substack{ \mathbf{u} \in U \cap (U \cap W)^\perp, \ \|\mathbf{u}\|_2 \leq 1 \\ \mathbf{w} \in W \cap (U \cap W)^\perp, \  \|\mathbf{w}\|_2 \leq 1 } \right\}.
\end{align*}
\end{definition}
Note that $\theta \in (0,\frac{\pi}{2}]$ implies $0 < \xi \leq 1$, and that $\xi(\mathcal{V}') \leq \xi(\mathcal{V})$ when $\mathcal{V}' \subseteq \mathcal{V}$.\footnote{We acknowledge the counter-intuitive property that $\theta =  \pi/2$ when $U \subseteq W$.}  
The constant $C_2$ in Lem.~\ref{MainLemma} is then:  
\begin{align}\label{Cdef2}
	C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ 1- \max_{\mathcal{G} \in {\mathcal{H} \choose r+1}} \xi( \bm{\mathcal{A}}_\mathcal{G} ) },
\end{align}
% C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \in {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1}} \xi( \{ \bm{\mathcal{A}}_S \}
which we remark is consistent with the assumption on $C_1$ in the proof of the case $k=1$ at the beginning of this section.

@@

It follows from $L_{2\mathcal{H}}(\mathbf{A}) > 0$ and the general linear position of the $\mathbf{x}_i$ that $L_k(\mathbf{AX}_{K}) > 0$; that is, the columns of the $n \times k$ matrix $\mathbf{AX}_K$ form a basis for $\bm{\mathcal{A}}_S$. Fixing $\mathbf{0} \neq \mathbf{y} \in \bm{\mathcal{A}}_S$, there then exists $\mathbf{0} \neq \mathbf{c} = (c_1, \ldots, c_k) \in \mathbb{R}^k$ such that $\mathbf{y} = \mathbf{AX}_K\mathbf{c}$. Setting \mbox{$\mathbf{\overline{y}} = \mathbf{B\overline{X}}_K\mathbf{c} \in \bm{\mathcal{B}}_{\overline S}$}, we have:
\begin{align*}
\|\mathbf{y} - \mathbf{\overline{y}}\|_2 
&= \|\sum_{i=1}^k c_i(\mathbf{AX}_K - \mathbf{B\overline{X}}_K)_i\|_2
\leq \varepsilon \sum_{i=1}^k |c_i| \\
&\leq \varepsilon \sqrt{k}  \|\mathbf{c}\|_2 
\leq \frac{\varepsilon}{L_k(\mathbf{AX}_K)} \|\mathbf{y}\|_2,
\end{align*}
where the last inequality follows directly from \eqref{Ldef}. From Def.~\ref{dDef}:
\begin{align}\label{rhs222}
d(\bm{\mathcal{A}}_S, \bm{\mathcal{B}}_{\overline S}) 
\leq \frac{\varepsilon}{  L_k(\mathbf{AX}_{K}) } \leq \varepsilon \frac{C_1}{C_2},
\end{align}
%
where the second inequality is due to $L_k(\mathbf{AX}_{K}) \geq L_k(\mathbf{AX}_{I(S)})$ and \eqref{Cdef1}. Finally, apply Lem.~\ref{MainLemma} with $\varepsilon < L_2(\mathbf{A})/C_1$. % L_2(A) > 0 by def of L, since L_2H(A) > 0 and H covers [m]
\end{proof}

@@

\begin{lemma}\label{DistanceToIntersectionLemma}
Fix $k \geq 2$. Let $\mathcal{V} = \{V_i\}_{i=1}^k$ be a collection of subspaces of $\mathbb{R}^m$ and set $V = \cap_{i = 1}^k V_i$. For  $\mathbf{x} \in \mathbb{R}^m$, we have (recalling \eqref{xi}):
\begin{align}\label{DTILeq}
\text{\rm dist}(\mathbf{x}, V) \leq \frac{1}{1 - \xi(\mathcal{V})} \sum_{i=1}^k \text{\rm dist}(\mathbf{x}, V_i).
\end{align}
\end{lemma}
\vspace{-.3 cm}
\begin{proof} 
Recall the orthogonal projection onto a subspace $V \subseteq \mathbb{R}^m$ is the mapping $\Pi_V: \mathbb{R}^m \to V$ that associates with each $\mathbf{x}$ its unique nearest point in $V$; i.e., $\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 = \text{\rm dist}(\mathbf{x}, V)$.
Next, observe:
\begin{align}\label{f}
\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 &\leq \|\mathbf{x} - \Pi_{V_k} \mathbf{x}\|_2 + \|\Pi_{V_k}  \mathbf{x} - \Pi_{V_k}\Pi_{V_{k-1}}\mathbf{x}\|_2 \nonumber \\
&\ \ \ + \cdots + \|\Pi_{V_k} \Pi_{V_{k-1}}\cdots \Pi_{V_1} \mathbf{x} - \Pi_V \mathbf{x}\|_2 \nonumber \\
&\leq \|\Pi_{V_k}\cdots\Pi_{V_{1}} \mathbf{x} - \Pi_V \mathbf{x}\|_2 + \sum_{\ell=1}^k \|\mathbf{x} - \Pi_{V_{\ell}} \mathbf{x}\|_2,
\end{align}
using the triangle inequality and that the spectral norm satisfies $\|\Pi_{V_{\ell}}\|_2 \leq 1$ for all $\ell$ (since $\Pi_{V_{\ell}}$ are orthogonal projections).

@@

Finally, substituting this into \eqref{f} and rearranging produces \eqref{DTILeq} after substituting $\xi(\mathcal{V})$ for $z$.
\end{proof}
\begin{lemma}\label{NonEmptyLemma} 
Fix an $r$-regular hypergraph $\mathcal{H} \subseteq 2^{[m]}$ satisfying the SIP. If the map $\pi: \mathcal{H} \to 2^{[\overline m]}$ has $\sum_{S \in \mathcal{H}} |\pi(S)| \geq \sum_{S \in \mathcal{H}} |S|$ and:
\begin{align}\label{cond}
	|\cap \pi(\mathcal{G})| \leq |\cap \mathcal{G} |,\ \ \   \text{for } \mathcal{G} \in {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1},
\end{align}
%
then $\overline m \geq m$, and if $\overline m < mr / (r-1)$ then the map $i \mapsto \cap \pi(\sigma(i))$ is an injective map to $[\overline m]$ from some $J \subseteq [m]$ of size $\overline m - r(\overline m - m)$. %In particular, if $\overline m = m$ then $\pi$ is induced by a permutation on $[m]$.
\end{lemma}
\begin{proof}
Consider the set: $T_1 := \{(i, S): i \in \pi(S), S \in \mathcal{H}\}$, which number $|T_1| = \sum_{S \in \mathcal{H}} |\pi(S)| \geq \sum_{S \in \mathcal{H}} |S| = \sum_{i \in [m]} \deg_\mathcal{H}(i) = mr$. Note that \eqref{cond} implies $|T_1| \leq \overline m r$, since otherwise pigeonholing the elements of $T_1$ with respect to their possible first indices $[\overline m]$ would imply $\deg_\mathcal{H}(i) > r$ for some $i \in [m]$. Thus, $\overline m \geq m$.

@@

We next show that \eqref{cond} holds.  Fix $\mathcal{G} \subseteq {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1}$. Since $\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})} \subseteq \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)}$, if $\cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)} = \{\textbf{0}\}$, then we must have $|\cap \pi(\mathcal{G})| = 0$ (as the columns of $\mathbf{B}_{\pi(S)}$ are linearly independent) and \eqref{cond} is trivially true. Suppose then that the intersection is not zero. Keeping in mind that $d(U',V) \leq d(U,V)$ whenever $U' \subseteq U$ and applying (in order) Lem.~\ref{spanIntersectionLemma}, Lem.~\ref{DistanceToIntersectionLemma}, \eqref{eqdim}, \eqref{GapUpperBound}, and \eqref{Cdef2} gives:
\begin{align}\label{randoml}
d(&\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})}, \bm{\mathcal{A}}_{\cap \mathcal{G}}  ) 
\leq d\left( \cap \bm{\mathcal{B}}_{\pi(\mathcal{G})}, \cap \bm{\mathcal{A}}_\mathcal{G} \right)
\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \cap \bm{\mathcal{B}}_{\pi(\mathcal{G})},\bm{\mathcal{A}_{T}} \right) }{ 1 - \xi( \bm{\mathcal{A}}_\mathcal{G} ) } \nonumber \\
&\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \bm{\mathcal{B}}_{\pi(T)},\bm{\mathbf{A}}_{T} \right) }{ 1 - \xi( \bm{\mathcal{A}}_\mathcal{G} ) }
\leq \frac{|\mathcal{G}| \varepsilon}{1 - \xi( \bm{\mathcal{A}}_\mathcal{G} )} 
\leq \frac{C_2 \varepsilon}{\max_j\|\mathbf{A}_j\|_2}. 
\end{align}
